---
title: "LA-HOLC-grade-inequality"
author: "Patty Park"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

### Data

#### EJScreen

We will be working with data from the United States Environmental
Protection Agency's EJScreen: Environmental Justice Screening and
Mapping Tool.

According to the [US EPA
website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:
> - to be more transparent about how we consider environmental justice
> in our work,
> - to assist our stakeholders in making informed decisions about
> pursuing environmental justice and,
> - to create a common starting point between the agency and the public
> when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
You will be working with block group data that has been downloaded from
the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To
understand the associated data columns, you will need to explore the
Technical Documentation and column description spreadsheet available in
the data folder. I also encourage you to explore the [limitations and
caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen)
of the data.

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

We will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed.

We will be working observations of birds from 2021 onward.

### Investigate the legacy of redlining in current environmental (in)justice

Before starting, we need to load any necessary packages in order to do our analysis and mapping

```{r }
library(tidyverse)
library(sf)
library(tmap)
library(spData)
library(gt)
```


Here I will read in the data from
```{r }
ejscreen <- st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")

lac <- ejscreen %>% 
  filter(CNTY_NAME == "Los Angeles County")
```

Make a map of wastewater discharge by census block groups. Indicate
which census block groups are above the 95th percentile of national
values for wastewater discharge by adding a centroid. (10 points)

```{r }
#REMEMBER COMPASS AND SCALE AND STUFF GRADICLES
# ASK ABOUT WHY SCALE BAR WONT MOVE
#ASK IF USING P_PWDIS OR PWDIS IS FOR WASTEWATER OF MAP (cause we are looking at it in percentile in the first place)

#filter for data that does not include an NA value
lac_simple_dis <- lac %>% filter(!is.na(P_PWDIS))


# create centroids for all PWDIS above the 95th percentile
lac_centroid_dis95 <- lac %>% #pip from lac
  select(P_PWDIS) %>% #select P_PWDIS
  filter(P_PWDIS > 95) %>% #filter for percentiles that are above the 95 percentiles
  st_centroid() #create a centroid for those filtered percentiles


lac_centroids_dis95_map <- tm_shape(lac_centroid_dis95) + #add the lac_centroid_dis95 data on top of first map
  tm_dots(fill = "red") #have them as dots, with color red

lac_simple_dis_map <- tm_shape(lac_simple_dis) + #create a map using the lac_simple dataset
  tm_graticules() + # add graticules on the map
  tm_fill(fill = 'P_PWDIS', #have fill be for P_PWDIS column
          palette = 'greens', #have the palette changed to green
          title = "Percentiles of Wastewater discharge") + #put title on legend
  tm_compass(type = "arrow", position = c("right", "top"), size = 3) + # add compass on map at the top right of the map
  tm_scale_bar(breaks = c(0, 15, 30, 45), position = c("left", "bottom")) + # scale bar is on map at the bottom right of the map (cannot move and manipulate)
  tm_layout(asp = 1.5, #change the aspect of the map
            title = "Wasterwater Discharge for Block Groups above 95 Percentile", #put a title on the map
            title.size = 2) +
   tm_credits("Red dots represent centroids of percentile wastewater discharge above 95",
             position = c("LEFT", "BOTTOM"))#change the font size of title

#combine two maps together and print combined maps
lac_simple_dis_map + lac_centroids_dis95_map
```

Find the percent of census block groups that have:
- less than 5% of the population is considered low income (5 points)

```{r }
#finding low income percent that is less than 5 percent
lowincpct_5 <- lac %>% #piping from lac
  select(LOWINCPCT) %>% #select LOWINCPCT
  filter(LOWINCPCT < 0.05) #filter for LOWINCPCT that is less than 5 percent

#filtering for just low income percent in a new dataframe
lac_lowincpct <- lac %>% #piping from lac
  select(LOWINCPCT) #select for just LOWINCPCT in a new dataframe

#divide the number of rows in lowincpct_5 by rows in lac_lowincpct and multiply by 100
(nrow(lowincpct_5) / nrow(lac_lowincpct)) *100
```

Find the percent of census block groups that are:
- above the 80th percentile for Particulate Matter 2.5 AND
- above the 80th percentile for Superfund proximity (10 points)

```{r }
#Filtering for particulate matter 2.5 and superfund proximity for above 80 percentile
lac_pm25_super_80 <- lac %>% #piping from lac
  filter(P_PM25 > 80 & P_PNPL > 80) #filter from P_PM25 and P_PNPL for above the 80 percentile

#divide number of rows in lac_pm25_super_80 by number of rows in lac
nrow(lac_pm25_super_80) / nrow(lac) * 100
```

Import redlining information for Los Angeles.

```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/downloads/geojson/CALosAngeles1939.geojson") %>%
  st_make_valid()

```

Make a map of historical redlining boundaries, colored by HOLC grade. (5
points)

```{r }
#print map that includes the historical redlining boundaries
tm_shape(LA_redlining) + #load LA_redlining data
  tm_graticules() + #print on gradicules on map
  tm_polygons(fill = "holc_grade", #have the fill on holc_grade column
              palette = c("green", "blue", "yellow", "red"), #change the palette to fit original holc grade color scheme
              breaks = c("A", "B", "C", "D")) +
  tm_shape(lac) + #load lac data
  tm_polygons(alpha = 0.5, #have transparancy be at 0.5
              lwd = 0.1) + #have line width at 0.1
    tm_compass(type = "arrow", position = c("right", "top"), size = 3) + # add compass on map at the top right of the map
  tm_scale_bar(position = c("left", "bottom")) + # scale bar is on map at the bottom right of the map (cannot move and manipulate)
  tm_layout(asp = 1.3, #change the aspect of the map
            title = "Los Angeles HOLC Grades on top of Los Angeles County", #put a title on the map
            title.size = 1) #change the font size of title

```

Find the number of census block groups that fall within areas with HOLC
grades hint: make sure the CRS match (15 points)

```{r eval=TRUE}
#check crs of Los Angeles County
st_crs(lac)

#transform the crs of LA_redlining to the crs of lac
LA_redlining_crs_t <- st_transform(LA_redlining, crs = st_crs(3857))

#join the two datasets by only filtering for the census block group that are completely within the HOLC grades
redlining_lac_join <- st_intersection(lac, LA_redlining_crs_t)

#print how many row there are in redlining_lac_join
nrow(redlining_lac_join)
```

Summarize current conditions based on EJScreen data within historical
redlining categories using the mean of the following variables:
-% low income.
- percentile for particulate Matter 2.5.
- percentile for low life expectancy.
- percentile for air toxics cancer risk (20 points)

```{r }
#find the means of the four variables we are looking for
redlining_mean_summary <- redlining_lac_join %>% #piping from redlining_lac_join
  group_by(holc_grade) %>% # grouping by holc_grade
  summarise(low_income_mean = mean(LOWINCPCT, na.rm = TRUE) * 100, #find the mean of LOWINCPCT
            pm25_mean = mean(P_PM25, na.rm = TRUE), #find the mean of P_PM25
            low_lifeexp_mean = mean(P_LIFEEXPPCT, na.rm = TRUE), #find the mean of P_LIFEEXPPCT
            cancer_mean = mean(P_CANCER, na.rm = TRUE)) %>% #find the mean of P_CANCER
  st_drop_geometry() %>% #get rid of geometry column
  gt() #use gt function to get nicer table

#print summary with means of variables we are interested in
redlining_mean_summary
```

Please write 2-3 sentences reflecting on these results. (5 points)
**Answer**: Overall, those in the A HOLC grade have a better standard of living than those in the D HOLC grade. I can assume that for those that have larger income, they have access to better health care and live in a cleaner area verses those that have lower income.

### Investigate the legacy of redlining in biodiversity observations

For bird observations from 2022 that fall within neighborhoods with HOLC
grads, find the percent of observations within each redlining categories
and plot results. hint: make sure that the bird observations have the
same CRS as redlining data. (20 points) 


```{r } 
#read in bird data
birds <- st_read("data/gbif-birds-LA") %>% filter(year == 2022)

#check the crs for LA_redlining
st_crs(LA_redlining) 

#transform crs and join dataframes
birds_crs_t <- st_transform(birds, crs = st_crs(4326)) #transform birds dataframe crs to LA_redlining crs
redlining_birds_join <- st_join(LA_redlining, birds_crs_t) #join both LA_redlining and new bird dataframe to get joined dataframe


redlining_bird_summary <- redlining_birds_join %>% #piping from redlining_birds_join
  group_by(holc_grade) %>% #group by holc grade
summarise(percentage = n()/nrow(redlining_birds_join)) #create a summary table by dividing each holc grade by the total rows in redlining_birds_join

#plot as bargraph
ggplot(data = redlining_bird_summary, #use redlining_bird_summary data
       aes(x = holc_grade, #have holc_grade be on x axis
           y = percentage)) + #have percentage be on y axis
  geom_bar(stat = "identity", #telling ggplot I want to aggregate manually
           aes(fill = holc_grade)) + #have fill be holc_grade
  scale_fill_manual(values=c("lightgreen", "lightskyblue", "yellow", "red"))+ #change color manually
  labs(x = "HOLC Grade", #change x label
  y = "Percentage (in decimal form)", #change y label
  title = "Percentage of bird sighting by HOLC Grade") #change title
```


Please write 2-3 sentences reflecting on these results. Are they
surprising? If so, why? What are some potential implications? (5 points)
**Answer**: I am surprised with the results I plotted, as HOLC Grade C and D have a higher percentage of bird sighting verses those in HOLC Grade A and B. I would have assumed the graph to have more bird sightings in HOLC Grade A and B rather than C and D. This could be that there are more resources found in C and D, such as humans throwing their trash away at the side of the street verses in a trash can. These birds would rather be around an area with constant food resources, which could be found in the C and D communities.
